Learning Auditory Similarities Between Instrumental Playing Techniques

### 2019-10-10 paper/paper.tex
write more on contributions

This article proposes a machine listening system which computes the dissimilarity in timbre between any two audio samples.
Our system composes two stages: unsupervised feature extraction and supervised metric learning.
The feature extraction stage is a nonlinear map which relies on the joint time--frequency scattering transform \cite{anden2015joint}, followed by per-feature Gaussianization \cite{lostanlen2018jasmp}.
It encodes patterns of spectrotemporal modulation in the query while offering numerical guarantees of stability to local deformations.
The metric learning stage is a linear map, optimized via large-margin nearest neighbors (LMNN) \cite{weinberger2009distance}.
It recombines scattering coefficients so as to fit a training set of human judgments \cite{mcadams1995psychres}.

The main originality of our model is that it encompasses a broad range of extended playing techniques, well beyond the so-called ``ordinary'' mode of acoustic production \cite{lostanlen2018extended}.
Specifically, we fit pairwise judgments for 78 different techniques arising from 16 instruments, some of which include removable timbre-altering devices such as mutes.
A second originality of our model is that we purposefully disregard the playing technique metadata underlying each audio sample during the training phase of our model.
Said otherwise, we rely on listeners, not performers, to define and evaluate the task at hand.
A comprehensive ablation study of our system demonstrates that all design choices presented hereafter benefit the retrieval precision of our system.

@@

On the other hand, the field of music psychology investigates timbre with the aim of discovering its physiological and behavioral foundations \cite{mcadams2009chapter}.
In this setting, prior knowledge, however accurate, of instrumentation does not suffice to conduct a study.
Rather, each excerpt must be played back to multiple human listeners.
Yet, collecting subjective responses to acoustic stimuli is a tedious and unscalable procedure, thus restricting the size of the musical corpus under study.
Such restriction hampers the applicability of iterative representation learning algorithms, such as stochastic gradient descent in artificial neural networks.
Nevertheless, advanced methods in electrophysiology allow to observe the firing patterns of biological neurons in the presence of controlled stimuli.
This observation, originally carried out on the ferret, has led to a comprehensive mapping of the primary auditory cortex, in terms of its spectrotemporal receptive fields (STRFs) \cite{depireux2001jneur}.
The STRF of a neuron is a function of time and frequency which represents the optimal predictor of its post-stimulus time histogram (PSTH) during exposure to a diverse range of auditory stimuli \cite{aertsen1981biolcyb}.
The simplest method to compute it in practice is by reverse correlation, i.e. by averaging all stimuli that trigger an action potential \cite{deboer1968biomed}.
Historically, STRFs were defined by their Wigner--Ville distribution \cite{flandrin1998book}, thereby sparing the choice of a tradeoff in time--frequency localization, but eliciting cross-term interferences \cite{eggermont1993hearing}.
Since then, the STRF of a neuron was redefined as a spectrographic representation of its spike-triggered average \cite{klein2000compneur}.
Although this new definition is necessarily tied to a choice of spectrogram parameters, it yields more interpretable patterns than a Wigner-Ville distribution.
In particular, a substantial portion of spectrographic STRFs exhibits a ripple-like response around a given region $(t, \lambda)$ of the time--frequency domain.
This response can be approximately described by a pair of scalar values: a temporal modulation rate $\alpha$ in Hertz and a frequential modulation rate $\beta$ in cycles per octave.
Interestingly, both $\alpha$ and $\beta$ appear to be arranged in a geometric series and independent from the selected center time $t$ and center frequency $\lambda$.
This observation has led auditory neuroscientists to formulate an idealized computational model for STRF, known as ``full cortical model'' \cite{chi2005jasa}, which densely covers the rate-scale domain $(\alpha, \beta)$.

@@

\begin{figure}
\caption{\csentence{STRF + scattering (upsampling STRF)}
A short description of the figure content should go here.}
\end{figure}

The reasons above make STRFs an appealing feature extractor for a perceptual description of timbral similarity across instrumental playing techniques.
Nonetheless, current implementations of STRF suffer from a lack of scalability, which explains why they have found few applications in MIR thus far.
Indeed, the full cortical model is usually computed via two-dimensional Fourier transforms over adjacent time--frequency regions, followed by averaging around specific rates and scales.
On the contrary, joint time--frequency scattering offers a faster extraction of spectrotemporal modulation while preserving properties of differentiability \cite{andreux2019jmlr} and invertibility \cite{lostanlen2019dafx}.
Such acceleration is made possible by discretizing the wavelet transforms involved in time--frequency scattering according to a multirate scheme, both along the time and the log-frequency variables.
This multirate scheme also has the benefit of drawing an explicit connection between scattering networks and deep convolutional networks, as both involves numerous convolutions with small kernels, pointwise rectifying nonlinearities, and dyadic pooling operations \cite{mallat2016philtrans}.

@@

%Our model borrows from both aforementioned lines of research in music information retrieval and music psychology.
%Indeed, the joint time--frequency scattering transform can be regarded as a deep convolutional network with fixed synaptic weights; or, alternatively, as an idealized computational model for STRFs \cite{anden2019joint}.
%We refer to \cite{bellet2015book} for a review of the state of the art in this domain.

@@

This representation does possess the desired invariance and stability properties, but its separable nature limits its power to adequately represent time--frequency structures that do not separate cleanly in time and frequency, such as chirps.
To remedy this, another approach was suggested by And\'{e}n et al. \cite{anden2015joint,anden2019joint} wherein the temporal wavelet decomposition of the scalogram is replaced with a two-dimensional, spectro-temporal decomposition.
Similarly to the cortical representation, the result is a four-dimensional array indexed by time, acoustic frequency, modulation frequency (or \emph{rate}), and spectral modulation frequency, known as \emph{quefrency} (or \emph{scale}).
As before, we compute the modulus and average in time, obtaining a set of coefficients known as second-order joint time--frequency scattering coefficients.

The definition of the scattering transform in terms of wavelet decompositions and modulus nonlinearities presents several advantages.
On the computational side, the critical bandwidth guarantees of wavelet transforms allows us to judiciously subsample the output, resulting in a lower-dimensional representation compared to the cortical representations.
The scattering transform also satisfies the aforementioned stability conditions, providing a guarantee that slightly deforming a signal only results in a negligible change in its scattering transform.
Furthermore, we may calculate the scattering tranforms of several model signals, including amplitude-modulated harmonic sounds and noise excitations \cite{anden2012scattering,anden2014deep}, frequency-modulated sinusoids (e.g., chirps) \cite{anden2012scattering,anden2019joint}, and beating phenomena \cite{anden2014deep}.
They have also enjoyed success in several audio classification and similarity retrieval tasks \cite{anden2011multiscale,bauge2013representing,anden2014deep,anden2019joint,lostanlen2018jasmp,lostanlen2018extended}.

@@

On the other hand, we note that the complexity of the learned model does not have to be high.
Indeed, for this task, a linear model on the scattering coefficients is sufficient.
There is no need for deep networks with large numbers of parameters to accurately represent the similarity information.
In other words, the scattering transform parametrizes the signal structure in a way that many relevant quantities (such as timbre similarity) can be extracted through a simple linear mapping.
This is in line with other classification results, where linear support vector machines applied to scattering coefficients have achieved significant success \cite{anden2015joint,anden2019joint}.

>>>>

This article proposes a machine listening system which computes the dissimilarity in timbre between any two audio samples.
Our system composes two stages: unsupervised feature extraction and supervised metric learning.
The feature extraction stage is a nonlinear map which relies on the joint time--frequency scattering transform \cite{anden2015mlsp}, followed by per-feature Gaussianization \cite{lostanlen2018jasmp}.
It encodes patterns of spectrotemporal modulation in the query while offering numerical guarantees of stability to local deformations.
The metric learning stage is a linear map, optimized via large-margin nearest neighbors (LMNN) \cite{weinberger2009distance}.
It recombines scattering coefficients so as to fit a training set of human judgments \cite{mcadams1995psychres}.

This article gathers three novel contributions.
First, our model encompasses a broad range of extended playing techniques, well beyond the so-called ``ordinary'' mode of acoustic production \cite{lostanlen2018extended}.
Specifically, we fit pairwise judgments for 78 different techniques arising from 16 instruments, some of which include removable timbre-altering devices such as mutes.
Secondly, we purposefully disregard the playing technique metadata underlying each audio sample during the training phase of our model.
Said otherwise, we rely on listeners, not performers, to define and evaluate the task at hand.
A comprehensive ablation study of our system demonstrates that all design choices presented hereafter benefit the retrieval precision of our system.
Thirdly, we supplement our quantitative benchmark with some visualizations of time--frequency scattering coefficients in the rate--scale domain, for various typical samples of realistic instrumental playing techniques.
These visualizations confirm the theoretical predictions of previous research \cite{anden2015mlsp}, which relied on simple synthetic sounds such as exponential chirps.
They are also in line with visualizations of the modulation power spectrum in auditory neurophysiology, while offering an accelerated algorithm for scalable feature extraction.

@@

On the other hand, the field of music psychology investigates timbre with the aim of discovering its physiological and behavioral foundations \cite{mcadams2009chapter}.
In this setting, prior knowledge, however accurate, of instrumentation does not suffice to conduct a study.
Rather, each excerpt must be played back to multiple human listeners.
Yet, collecting subjective responses to acoustic stimuli is a tedious and unscalable procedure, thus restricting the size of the musical corpus under study.
Such restriction hampers the applicability of iterative representation learning algorithms, such as stochastic gradient descent in artificial neural networks.
Nevertheless, advanced methods in electrophysiology allow to observe the firing patterns of biological neurons in the presence of controlled stimuli.
This observation, originally carried out on the ferret, has led to a comprehensive mapping of the primary auditory cortex, in terms of its spectrotemporal receptive fields (STRFs) \cite{depireux2001jneur}.
The STRF of a neuron is a function of time and frequency which represents the optimal predictor of its post-stimulus time histogram (PSTH) during exposure to a diverse range of auditory stimuli \cite{aertsen1981biolcyb}.
The simplest method to compute it in practice is by reverse correlation, i.e. by averaging all stimuli that trigger an action potential \cite{deboer1968biomed}.
Historically, STRFs were defined by their Wigner--Ville distribution \cite{flandrin1998book}, thereby sparing the choice of a tradeoff in time--frequency localization, but eliciting cross-term interferences \cite{eggermont1993hearing}.
Since then, the STRF of a neuron was redefined as a spectrographic representation of its spike-triggered average \cite{klein2000compneur}.
Although this new definition is necessarily tied to a choice of spectrogram parameters, it yields more interpretable patterns than a Wigner-Ville distribution.
In particular, a substantial portion of spectrographic STRFs exhibits a ripple-like response around a given region $(t, \lambda)$ of the time--frequency domain.
This response can be approximately described by a pair of scalar values: a temporal modulation rate $\alpha$ in Hertz and a frequential modulation rate $\beta$ in cycles per octave.
Interestingly, both $\alpha$ and $\beta$ appear to be arranged in a geometric series and independent from the selected center time $t$ and center frequency $\lambda$.
This observation has led auditory neuroscientists to formulate an idealized computational model for STRF, known as ``full cortical model'' \cite{chi2005jasa}, which densely covers the rate--scale domain $(\alpha, \beta)$.

@@

%\begin{figure}
%\caption{\csentence{STRF + scattering (upsampling STRF)}
%A short description of the figure content should go here.}
%\end{figure}

The reasons above make STRFs an appealing feature extractor for a perceptual description of timbral similarity across instrumental playing techniques.
Nonetheless, current implementations of STRF suffer from a lack of scalability, which explains why they have found few applications in MIR thus far.
Indeed, the full cortical model is usually computed via two-dimensional Fourier transforms over adjacent time--frequency regions, followed by averaging around specific rates and scales.
On the contrary, joint time--frequency scattering offers a faster extraction of spectrotemporal modulations while preserving properties of differentiability \cite{andreux2019jmlr} and invertibility \cite{lostanlen2019dafx}.
Such acceleration is made possible by discretizing the wavelet transforms involved in time--frequency scattering according to a multirate scheme, both along the time and the log-frequency variables \cite{anden2019tsp}.
This multirate scheme also has the benefit of drawing an explicit connection between scattering networks and deep convolutional networks, as both involves numerous convolutions with small kernels, pointwise rectifying nonlinearities, and dyadic pooling operations \cite{mallat2016philtrans}.
Moreover, a quantitative benchmark over Medley-solos-DB has demonstrated that joint time--frequency scattering, unlike purely temporal scattering, outperforms deep convolutional networks in supervised musical instrument classification, even in a relatively large data regime with 500 to 5k samples per class.
Recently, a previous publication

@@

%Our model borrows from both aforementioned lines of research in music information retrieval and music psychology.
%Indeed, the joint time--frequency scattering transform can be regarded as a deep convolutional network with fixed synaptic weights; or, alternatively, as an idealized computational model for STRFs \cite{anden2019tsp}.
%We refer to \cite{bellet2015book} for a review of the state of the art in this domain.

@@

This representation does possess the desired invariance and stability properties, but its separable nature limits its power to adequately represent time--frequency structures that do not separate cleanly in time and frequency, such as chirps.
To remedy this, another approach was suggested by And\'{e}n et al. \cite{anden2015mlsp,anden2019tsp} wherein the temporal wavelet decomposition of the scalogram is replaced with a two-dimensional, spectro-temporal decomposition.
Similarly to the cortical representation, the result is a four-dimensional array indexed by time, acoustic frequency, modulation frequency (or \emph{rate}), and spectral modulation frequency, known as \emph{quefrency} (or \emph{scale}).
As before, we compute the modulus and average in time, obtaining a set of coefficients known as second-order joint time--frequency scattering coefficients.

The definition of the scattering transform in terms of wavelet decompositions and modulus nonlinearities presents several advantages.
On the computational side, the critical bandwidth guarantees of wavelet transforms allows us to judiciously subsample the output, resulting in a lower-dimensional representation compared to the cortical representations.
The scattering transform also satisfies the aforementioned stability conditions, providing a guarantee that slightly deforming a signal only results in a negligible change in its scattering transform.
Furthermore, we may calculate the scattering tranforms of several model signals, including amplitude-modulated harmonic sounds and noise excitations \cite{anden2012scattering,anden2014deep}, frequency-modulated sinusoids (e.g., chirps) \cite{anden2012scattering,anden2019tsp}, and beating phenomena \cite{anden2014deep}.
They have also enjoyed success in several audio classification and similarity retrieval tasks \cite{anden2011multiscale,bauge2013representing,anden2014deep,anden2019tsp,lostanlen2018jasmp,lostanlen2018extended}.

@@

On the other hand, we note that the complexity of the learned model does not have to be high.
Indeed, for this task, a linear model on the scattering coefficients is sufficient.
There is no need for deep networks with large numbers of parameters to accurately represent the similarity information.
In other words, the scattering transform parametrizes the signal structure in a way that many relevant quantities (such as timbre similarity) can be extracted through a simple linear mapping.
This is in line with other classification results, where linear support vector machines applied to scattering coefficients have achieved significant success \cite{anden2015mlsp,anden2019tsp}.
